Modal Abstractions for Virtualizing Memory Addresses

### 2023-02-21 programlogic.tex
+

\begin{itemize}
\item  After a series of masking imposed on the virtual memory address (e.g. \textsf{vaddr} 64bits address [0-64]) :
  \begin{itemize}
  \item Page Map Level-4 Translation (L4): Calculating the offset to the 4 \textsf{Kbyte Physical Page} table to obtain the  physical address by  which resides in the last 12 bits of the virtual memory address, $\locsf|^{12}$,  shown in Figure \ref{}.  
  \item Page Directory Pointer Level Translation (L3):
  \item Page Directory Table Level Translation (L2):  we obtain the address of the page table via \textsf{PDE}$|^{52}$ and picking the offset from the virtual memory address via imposing the mask to obtain page table offset bits (9 bits [20-12] shown in Figure \ref{}) to obtain the page table entry (\textsf{PTE}).
  \item Page Table Level Translation (L1): 
  \end{itemize}
 \item then after obtainini the physical page address, we obtain the value 
\end{itemize}

\paragraph{Virtual-Memory points-to} As we see in Figure \fref{fig:}, virtual address mapping is obtained via traversal of page table, i.e. indirection provided by physical memory lookups. As expected, we define the virtual-points to relation, ($\pfpointsto\vaddr\locsf\qfrac\vpts$), in terms of multiple physical memory mappings representing the indirection shown in Figure \fref{fig:}
\begin{figure*}
\[
\begin{array}{l}
  \ppointsto\vaddr\locsf\vpts \stackrel{def}{=} \lambda \tlf\Loc.\\
  \exists_{\tlfoff\Locn \;, \tltoff\Locn \;, \tltwoff\Locn \;,\tlooff\Locn \;, \tpgoff\Loctw \;, \tlt\Loc \;, \tltw\Loc \;, \tlo\Loc \tpg\Loc} \ldotp \\
  \ulcorner \textsf{aligned } \vaddr \urcorner \star 
   \ulcorner \vaddr = \locsx :: \lfoff :: \ltoff :: 
   \ltwoff :: \looff :: \pgoff \urcorner \star\\
  \pfpointsto{\lvlsum\crt\lfoff}{\lvlbor\lt}\qfracfotsss\ppts \star 
  \pfpointsto{\lvlsum\lt\ltoff}{\lvlbor\ltw}\qfracfotss\ppts \star \\
  \pfpointsto{\lvlsum\ltw\ltwoff}{\lvlbor\lo}\qfracfots\ppts \star 
  \pfpointsto{\lvlsum\lo\looff}{\lvlbor\pg}\qfracfot\ppts \star \\
  \ppointsto{\pageptstosum\pg\pgoff}\locsf\ppts 
\end{array}
\]
\caption{Virtual Points-to Relation}
\todo[inline]{These fractions aren't quite right (though I see you added the variance between levels), I can walk you through in our meeting tomorrow.}
  \label{fig:virtualpointsto}
\end{figure*}
\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.

>>>>

\begin{itemize}
\item  After a series of masking imposed on the virtual memory address (e.g. \textsf{vaddr} 64bits address [0-64]) :
  \begin{itemize}
  \item Page Map Level-4 Translation (L4): Calculating the offset to the 4 \textsf{Kbyte Physical Page} table to obtain the  physical address by  which resides in the last 12 bits of the virtual memory address, $\locsf|^{12}$,  shown in Figure \ref{}.
    \[  \pfpointsto{\lvlsum\crt\lfoff}{\lvlbor\lt}\qfracfotsss\ppts \star \]
  \item Page Directory Pointer Level Translation (L3):
    \[ \pfpointsto{\lvlsum\lt\ltoff}{\lvlbor\ltw}\qfracfotss\ppts \star \]
  \item Page Directory Table Level Translation (L2):  we obtain the address of the page table via \textsf{PDE}$|^{52}$ and picking the offset from the virtual memory address via imposing the mask to obtain page table offset bits (9 bits [20-12] shown in Figure \ref{}) to obtain the page table entry (\textsf{PTE}).
    \[  \pfpointsto{\lvlsum\ltw\ltwoff}{\lvlbor\lo}\qfracfots\ppts \star \]
  \item Page Table Level Translation (L1):
    \[  \pfpointsto{\lvlsum\lo\looff}{\lvlbor\pg}\qfracfot\ppts \star \]
  \end{itemize}
\item then after obtainini the physical page address, we obtain the value
  \[  \ppointsto{\pageptstosum\pg\pgoff}\locsf\ppts \]
\end{itemize}

\subsection{Aliasing/Sharing Physical Pages}
As we see in Figure \fref{fig:}, virtual address mapping is obtained simply in two fold: 1. traversing the page tables to reach the page table entry 2. fetch the physical address and the value it is adressing (Section \ref{fig:physicalpointsto}). This two fold access pattern enabling sharing of pages by virtual memory addresses exhibits logically sound representation against virtual memory operations some of which ((e.g. mapping a new page) we explain in detail in Section \ref{}.

The very first assurance our logical constructions need to be able assert is the \textit{existence} of the page table walk. A naive attempt to assert the existence of the page-walk for a certain virtual memory translation would look like the one in Figure \ref{figure:strongvirtualpointsto}, in which we see that a virtual address with a certain fractional permissions at each level of the page-table walk gives us sound assurance of the path existence reaching to a page table entry, consequently, with a full-ownership of page address, we can obtain and update the value mapped value at this physical page address through our virtual memory address. 
\begin{figure*}
\[
\begin{array}{l}
  \ppointsto\vaddr\locsf\vpts \stackrel{def}{=} \lambda \tlf\Loc.\\
  \exists_{\tlfoff\Locn \;, \tltoff\Locn \;, \tltwoff\Locn \;,\tlooff\Locn \;, \tpgoff\Loctw \;, \tlt\Loc \;, \tltw\Loc \;, \tlo\Loc \tpg\Loc} \ldotp \\
  \ulcorner \textsf{aligned } \vaddr \urcorner \star 
   \ulcorner \vaddr = \locsx :: \lfoff :: \ltoff :: 
   \ltwoff :: \looff :: \pgoff \urcorner \star\\
  \pfpointsto{\lvlsum\crt\lfoff}{\lvlbor\lt}\qfracfotsss\ppts \star 
  \pfpointsto{\lvlsum\lt\ltoff}{\lvlbor\ltw}\qfracfotss\ppts \star \\
  \pfpointsto{\lvlsum\ltw\ltwoff}{\lvlbor\lo}\qfracfots\ppts \star 
  \pfpointsto{\lvlsum\lo\looff}{\lvlbor\pg}\qfracfot\ppts \star \\
  \ppointsto{\pageptstosum\pg\pgoff}\locsf\ppts 
\end{array}
\]
\caption{A Strong Virtual Points-to Relation}
\todo[inline]{These fractions aren't quite right (though I see you added the variance between levels), I can walk you through in our meeting tomorrow.}
  \label{fig:strongvirtualpointsto}
\end{figure*}
One might have already observed that this virtual points-to definition is too strong to specify any virtual memory operations, some of which we explain within as part of experiement in Section \ref{}, require reasoning of sharing of physicall pages accross virtual memory addresses because this strong virtual points-to relation doesn’t care about owning those fractional resources, it only cares that something ensures the correct page table walk exists. Therefore we need to weaken the virtual-pointsto to capture the acutal sharing accross the virtual addresses.

\subsection{Address-Spaces as a Modality}
\subsubsection{Modal Resource Context}
\subsubsection{Interaction: address-space switch as a modal-entailment}

\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.
